Supplementary Information

### 2020-08-23 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

**Here, we develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
(4) Are interactions between DBH and climate drivers common?

@@

*Testing for climate-DBH interactions*

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers. Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model). Plotted are responses of species for which $PET$ was included in the top model. (THIS IS STILL CONFUSING--maybe putting box around months identified by climwin)](tables_figures/quilt_comparison.png)

@@

Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
Responses were most commonly positive, and were most pronounced at the driest site (Little Tesque). 
Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

Temperature responses were included in the best model at all sites *(but BCI? [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations.
Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this? [ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

@@

**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. 4).** 
While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. 4).
On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. 4).
At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.

@@

*Climate-DBH interactions*

 (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)

@@

* nonlinear effects occur sometimes

@@

DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. DEN is new alternative.

*Climate-DBH interactions*

>>>>

**Here, we develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
(4) How do climate drivers additively and interactively shape growth?

@@

*Additive and interactive effects of climate and DBH*

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers. Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model). Plotted are responses of species for which $PET$ was included in the top model. (THIS IS STILL CONFUSING-- add months prior in (a), maybe put box around months identified by climwin)](tables_figures/quilt_comparison.png)

@@

Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
Responses were most commonly positive, and were most pronounced at the driest site (Little Tesuque). 
Precipitation terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
In some cases, the non-linearity was quite pronounced (e.g., Little Tesuque), with the most common pattern (#%) being a decelerating increase.

Temperature responses were included in the best model at all sites *(but BCI? [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations.
Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Temperature terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
The non-linearity was rarely pronounced...

@@

**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. 4).** 
While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. 4).
On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. 4).
At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.

@@

*Additive and interactive effects of climate and DBH*

**Both DBH and climate were included in the majority of top models, regardless of the growth metric used.** 
Specifically, $DBH$ was included in #% of top models for $\Delta r$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
In general, $DBH$ explained more variation in growth rates than did climate (Fig. 5).

**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. 5, S#_#).** (*It will be easier to fill this in once we complete [ISSUE #66 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/66)*).
The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. 5).
However, there were exceptions (e.g., at Little Tesuque; Fig. 5).
The relative importance of climate was modest at sites including SCBI, ..., and stronger at sites including Little Tesuque, ... (Figs 5, S#-S#).

![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are modest (SCBI, left panel) and strong (Little Tesuque, right panel).**](tables_figures/full_model_comparison.png)

@@

* nonlinear effects are prevalent

@@

*Additive and interactive effects of climate and DBH*

**We show that DBH is a strong driver of growth (any metric) relative to climate and that climate sensitivity often varies with DBH-- findings that have important implications for understanding changes in forest productivity in response to climate change.**
The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity (REFS; McDowell et al. 2020).
The prevalence of interactive $DBH$-climate effects reinforces the concepts that drought often has a disproportionate impact on large trees (REFS) and that *temperature sensitivity can vary across size classes (Rollinson et al. in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.
However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.
Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity (REFS), but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
Further work will be required to combine this model with forest census data in order to understand the climate sensitivity of whole-forest productivity.

All this points to the potential importance of considering DBH in climate reconstructions.
Traditional dendro methods ignore DBH.
DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. DEN is new alternative.

### 2020-08-23 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE, caption = "Table S1. List of species analyzed", escape = F)%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

@@

>>>>

## Appendix S3. Dealing with rapidly changing climate and tree growth

[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25)

Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesuque, New Mexico, USA (LT).
At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].

@@

![**Figure S2 | Example comparison of climwin output when the optimal climate variable and time window were identical across growth metrics.** Example is temperature variable group at Little Tesuque. Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png)

![**Figure S3 | Example comparison of climwin output when the optimal climate variable differed across growth metrics, but the selected time window was similar.** Example is precipitation variable group at Little Tesuque. Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)

![**Figure S4 | Example comparison of climwin output when both the optimal climate variable and selected time window differed across growth metrics.** Example is precipitation variable group at Harvard Forest. Here, *climwin* identified precipitation frequency (WET) as the strongest climate variable for $\Delta r$, but precipitation amount (PRE) as the strongest climate variable for $BAI$ and $\Delta AGB$. The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)
